%% Needed for the postscript graphics.
%%\input epsf

%% My macros.
\input macro

%% The length of a page is 10in.
%%\vsize=10in

%\font\helv = phvr
%\font\times = ptmr
%\font\zap = pzcmi at 12pt
%\font\pal = pplro at 12pt

%% Don't indent paragraphs.
\parindent=0pt
\parskip=.5\baselineskip
\baselineskip = 1.1\baselineskip

\footline{\hss\tenrm\folio\hss}

\mansubtitle{Measure Theoretic Conditional Expectation}
        {in an Elementary Setting}
{R. Scott McIntire}{May 2, 2026}

\section{Overview}
The measure theoretic approach to conditional expectation can be confusing
when compared to the traditional approach.
In what follows we go through a conditional 
expectation problem within a discrete and familiar setting in an attempt 
to reduce this confusion. One of the most confusing/disturbing ideas that 
one must confront from the very definition of conditional expectation is 
the notion of non-measurable functions. We work through an explicit
example where we demonstrate that a given function is not measurable in 
the ``conditional'' $\sigma$-algebra.

We can think of the conditional expectation of a function, $f$, on a set, $\Lambda$,
as a weighted average of $f$ over $\Lambda$.

From this perspective, you can think of the conditional expectation as
giving the ``best'' representation of $f$ with respect to a coarse $\sigma$-algebra,
given that $f$ lives on a finer one. Just as the ``best'' representation of an
image at a larger pixel scale (coarse mesh) would be an average over the smaller
scale pixels (refined mesh) that make up the larger pixel.
 

\section{Elementary Probability Example}
Let us start with a simple example. We have a die with six faces, $D_1, D_2, D_3, D_4, D_5, D_6$.
Any roll of the die will produce one of these faces with an equal probability. Let's further suppose
that the value of each side is the number of dots on the die. Now suppose that when the die is
rolled, we don't see the die, but someone tells us partial information -- only that the die is ``even'' or ``odd''.

{\bf Question:\/} Given that we only know the even or oddness of the die, what is the ``best'' guess of the die value?
Let's try to answer this question using the measure theoretic definition of conditional expectation.

Let $X = \{D_1, D_2, D_3, D_4, D_5, D_6\}$ and define a function $P$ by 
$P(D_i) = \frac{1}{6}$, for $i\in \{1,2,3,4,5,6\}$. The intent is that $P$ will 
become a probability measure for the space we construct.
Let ${\cal E} = 2^X$ be the $\sigma$-algebra 
consisting of the power set of $X$. We extend $P$ for every element in the $\sigma$-algebra.
Since the $\sigma$-algebra consists of all sets 
we need an assignment for an arbitrary set, $A$. 
The assignment is $P(A) = \frac{|A|}{6}$; that is, the cardinality of the set divided by 6.
We now have a measure space; in fact, a probability space: $(X,{\cal E},P)$.
Note that for a probability space we need a sample space, $X$, a $\sigma$-algebra of sets
(in the discrete case just an algebra) -- whose elements are the {\it events\/} -- and a function $P$ which takes
elements of the $\sigma$-algebra to [0,1] with the following properties:

\beginEnum
\item{$P(X) = 1$.}
\item{$P(\emptyset) = 0$.}
\item{$
P\bigl(\mathop{\bigcup}_{i=1}^\infty A_i\bigr) = \sum_{i=1}^\infty P(A_i)
\quad {\rm when}\;  A_k \cap A_j = \emptyset \quad k\neq j
$;}
\endEnum

We now consider a random variable from which we will get a sub $\sigma$-algebra.
Let 
$$
g(D_i) = \cases{0 & if $i$ is even;\cr
1 & if $i$ is odd. \cr
}
$$
In general, the $\sigma$-algebra generated from $g$ is the algebra generated by preimages
of Borel sets in $(-\infty,\infty)$. In the discrete case this reduces to the algebra of sets generated from
$g^{-1}(a)$, $a \in (-\infty, \infty)$. Clearly, the interesting sets come
from the values $0$ and $1$, all other values lead to the empty set. Consequently,
$$
\eqalignno{
	{\cal F} & = \{\emptyset, \{D_1,D_3,D_5\}, \{D_2,D_4,D_6\}, \{D_1,D_2,D_3,D_4,D_5,D_6\}\} & \cr
}
$$
In a discrete space a function, $f$, is measurable with respect to a $\sigma$-algebra if 
$f^{-1}(a)$ is an element in the $\sigma$-algebra for all $a\in (-\infty, \infty)$.
This has implications for the $\sigma$-algebra ${\cal F}$. 

{\bf Claim:\/} Any function, $f$,  
which is measurable over ${\cal F}$ has the property that $f$ is constant on the sets 
$\{D_1,D_3,D_5\}$ and $\{D_2,D_4,D_6\}$. More generally, we claim that any 
measurable function, $f$, in a given $\sigma$-algebra must be constant on 
the {\it minimal\/} elements of the algebra --  elements which have no non-trivial subsets.%
\footnote{\kern 0.5pt \raise 0.5ex \hbox{\dag}}{%
	Specifically, in a discrete setting, a set $Z$ in a $\sigma$-algebra, 
	${\cal H}$, is {\it minimal\/} in ${\cal H}$ if 
there is no non-empty, strict subset, $Y$, of $Z$ with $Y \in {\cal H}$.}

To see this suppose that $f(D_1)$ differs from $f(D_3)$.
Then the set, $A = \{D_1, D_3, D_5\} \cap f^{-1}(f(D_1))$, is a {\it strict\/}, non-trivial subset of 
$\{D_1, D_3, D_5\}$. This is true since $A$, by construction, is a subset of $\{D_1, D_3, D_5\}$;
$A$ contains $D_1$; and $A$ does not contain $D_3$.
Since ${\cal F}$ is a $\sigma$-algebra and $A$ is the intersection of the ${\cal F}$-measurable sets
$\{D_1, D_3, D_5\}$ and $f^{-1}(f(D_1))$, $A$ must be in ${\cal F}$.
However, we know that $\{D_1, D_3, D_5\}$ has no strict non-trivial subset in ${\cal F}$  -- contradiction. 
Therefore, our premise that $f(D_1)$ and $f(D_3)$ could 
take differing values is incorrect.  The same argument shows that $f(D_1)$ and 
$f(D_5)$ do not differ.
One can repeat the above argument to show that
$f$ is constant on the other minimal set $\{D_2,D_4,D_6\}$.

Notice that while any function over the measure space $(X,{\cal E},P)$ is measurable, we 
can write down a specific function that is non-measurable with respect to ${\cal F}$.
We know that all we have to do is come up with a function that differs on either of the 
sets $\{D_1, D_3, D_5\}$ or $\{D_2, D_4, D_6\}$. For instance, the function:  
$f(D_i) = i$, for $i\in\{1,2,3,4,5,6\}$, 
is a non-measurable function in ${\cal F}$.

\section{Conditional Expectation}
Given a probability space $(X,{\cal E},P)$, the conditional expectation of an
${\cal E}$-measurable, $P$-integrable function $f$ with respect to a sub $\sigma$-algebra ${\cal F}$ is the {\it unique\/} ${\cal F}$-{\it measurable\/}
function labeled, $\E{f \mid {\cal F}}$, such that%
\footnote{\kern 0.5pt \raise 0.5ex \hbox{\ddag}}{%
That such a unique function exists is a consequence of the Radon-Nikodym theorem.}
$$
\eqalignno{
\int_{\Lambda} \E{f \mid {\cal F}} \, dP & 
= \int_{\Lambda} f \, dP \quad \forall \Lambda \in {\cal F} & (1a) \cr
}
$$
Let us write this again in, perhaps, an unusual way.
$$
\eqalignno{
	\int_{\Lambda} \E{f \mid {\cal F}} \, dP_{\cal F} & 
= \int_{\Lambda} f \, dP \quad \forall \Lambda \in {\cal F} & (1b) \cr
}
$$
Equation (1b) is the same equation as (1a), but with the measure on the left written as
$P_{\cal F}$ -- the restriction of $P$ to the sub $\sigma$-algebra ${\cal F}$. As a number,
$P_{\cal F}(\Lambda) = P(\Lambda)$ for every $\Lambda \in {\cal F}$; the subscript is a
reminder that the integrand on the left is required to be ${\cal F}$-measurable for the
left-hand integral to be well-defined.

Although it seems that $f$ -- itself -- satisfies (1a), we have
to be careful. The right-hand side of (1b) integrates an ${\cal E}$-measurable function
against $P$, while the left-hand side requires an ${\cal F}$-measurable integrand against
$P_{\cal F}$. Consequently, while $f$ seems like a natural candidate for the conditional
expectation, it is not necessarily ${\cal F}$-measurable. Typically, $f$ is
{\it non-measurable\/} with respect to the sub $\sigma$-algebra, ${\cal F}$.
However, if $f$ is ${\cal F}$-measurable then, by the definition above,
it is its own conditional expectation.

The modern theory of probability -- including conditional expectation -- 
relies on measure theory which has its roots in Lebesgue measure theory -- 
used to provide an alternative to the theory of Riemann Integration.
From Lebesgue Measure Theory one is introduced to measurable and non-measurable sets/functions. 
The problem, in the context of Lebesgue, is that non-measurable sets/functions exist but 
specific examples are hard to come by. However, non-measurable functions appear 
{\it implicitly\/} in the definition of conditional expectation 
in that if they didn't the definition wouldn't be of any interest.

If one started out learning measure theory from the Lebesgue theory, the 
definition of conditional expectation might make readers somewhat uneasy as 
we are confronted with non-measurable sets from the very start.
From this context, it is harder to get an intuitive idea of conditional expectation. 

In the next section we compute the measure theoretic conditional expectation 
of a discrete function. 
In the process, we provide an intuitive idea of the 
measure-theoretic definition of conditional expectation which coincides with 
the traditional approach -- except in the ``singular'' case
which we discuss in the ``Redux'' section of this paper.

\section{Example Calculation of Conditional Expectation}
Consider the function from the second section on an elementary die example, $f(D_i) = i$,
which is measurable in the space $(X,{\cal E},P)$. Let ${\cal F}$ be the 
sub $\sigma$-algebra of that section.
We now compute the conditional expectation of $f$ with respect to ${\cal F}$.
Using (1b) we choose the {\it minimal\/} sets: $\Lambda_1 = \{D_1,D_3,D_5\}$ 
and $\Lambda_2 = \{D_2,D_4,D_6\}$. Since the conditional expectation function is constant
on each of these sets, equation (1b) will provide a way to find the value on each set.
Since the union of $\Lambda_1$ and $\Lambda_2$ constitutes the entire set, $X$, we will know
the value of the conditional expectation function on all of $X$; hence we will know the
conditional expectation function.

Proceeding, we have
$$
\eqalignno{
	\int_{\Lambda_1} \E{f \mid {\cal F}} \, dP_{\cal F} & 
	= \int_{\Lambda_1} f \, dP & (2) \cr
}
$$
and
$$
\eqalignno{
	\int_{\Lambda_2} \E{f \mid {\cal F}} \, dP_{\cal F} & 
	= \int_{\Lambda_2} f \, dP & (3) \cr
}
$$
We know $\E{f \mid {\cal F}}$ is {\it constant\/} on $\Lambda_1$. We label this value as
$\E{f \mid {\cal F}}\!(\Lambda_1)$. From (2) we have 

$$
\eqalignno{
\int_{\Lambda_1} \E{f \mid {\cal F}} \, dP_{\cal F}  & =  \int_{\Lambda_1} f \, dP & \cr
\E{f \mid {\cal F}}\!(\Lambda_1) \int_{\Lambda_1} \, dP_{\cal F} & = \int_{\Lambda_1} f \, dP & \cr 
\E{f \mid {\cal F}}\!(\Lambda_1) \cdot P_{\cal F}(\Lambda_1) & =  \int_{\Lambda_1} f \, dP = f(D_1) \cdot P(D_1) + f(D_3) \cdot P(D_3) + f(D_5) \cdot P(D_5) & (**) \cr 
\E{f \mid {\cal F}}\!(\Lambda_1)  & =  \frac{\int_{\Lambda_1} f \, dP}{P_{\cal F}(\Lambda_1)} = f(D_1) \cdot \frac{P(D_1)}{P(\Lambda_1)} + f(D_3) \cdot \frac{P(D_3)}{P(\Lambda_1)} + f(D_5) \cdot \frac{P(D_5)}{P(\Lambda_1)} & \cr 
\E{f \mid {\cal F}}\!(\Lambda_1)  & =  \frac{\int_{\Lambda_1} f \, dP}{P_{\cal F}(\Lambda_1)} = 1 \cdot \frac{\frac{1}{6}}{\frac{1}{2}} + 3 \cdot \frac{\frac{1}{6}}{\frac{1}{2}} + 5 \cdot \frac{\frac{1}{6}}{\frac{1}{2}} & \cr 
\E{f \mid {\cal F}}\!(\Lambda_1)  & =  \frac{\int_{\Lambda_1} f \, dP}{P_{\cal F}(\Lambda_1)} = 1 \cdot \frac{1}{3} + 3 \cdot \frac{1}{3} + 5 \cdot \frac{1}{3} & \cr 
\E{f \mid {\cal F}}\!(\Lambda_1)  & = 3 & \cr
}
$$

Likewise, $\E{f \mid {\cal F}}$ is constant on the set $\Lambda_2$. 
As we did above, we find the value of $\E{f \mid {\cal F}}$ on the set $\Lambda_2$.
As with $\Lambda_1$, label $\E{f \mid {\cal F}}\!(\Lambda_2)$ as the constant value of $\E{f \mid {\cal F}}$ 
on $\Lambda_2$.
We have

$$
\eqalignno{
\int_{\Lambda_2} \E{f \mid {\cal F}} \, dP_{\cal F}  & =  \int_{\Lambda_2} f \, dP & \cr
\E{f \mid {\cal F}}\!(\Lambda_2) \int_{\Lambda_2} \, dP_{\cal F} & = \int_{\Lambda_2} f \, dP & \cr 
\E{f \mid {\cal F}}\!(\Lambda_2) \cdot P_{\cal F}(\Lambda_2) & =  \int_{\Lambda_2} f \, dP = f(D_2) \cdot P(D_2) + f(D_4) \cdot P(D_4) + f(D_6) \cdot P(D_6) & \cr 
\E{f \mid {\cal F}}\!(\Lambda_2)  & =  \frac{\int_{\Lambda_2} f \, dP}{P_{\cal F}(\Lambda_2)} = f(D_2) \cdot \frac{P(D_2)}{P(\Lambda_2)} + f(D_4) \cdot \frac{P(D_4)}{P(\Lambda_2)} + f(D_6) \cdot \frac{P(D_6)}{P(\Lambda_2)} & \cr
\E{f \mid {\cal F}}\!(\Lambda_2)  & =  \frac{\int_{\Lambda_2} f \, dP}{P_{\cal F}(\Lambda_2)} = 2 \cdot \frac{\frac{1}{6}}{\frac{1}{2}} + 4 \cdot \frac{\frac{1}{6}}{\frac{1}{2}} + 6 \cdot \frac{\frac{1}{6}}{\frac{1}{2}}  & \cr
\E{f \mid {\cal F}}\!(\Lambda_2)  & =  \frac{\int_{\Lambda_2} f \, dP}{P_{\cal F}(\Lambda_2)} = 2 \cdot \frac{1}{3} + 4 \cdot \frac{1}{3} + 6 \cdot \frac{1}{3} & \cr
\E{f \mid {\cal F}}\!(\Lambda_2)  & = 4 & \cr
}
$$
Since a function is determined once we know what its values are on every $x \in X$, 
we have found the conditional expectation function, $\E{f \mid {\cal F}}$, as we know 
its values on every $x \in X$.\footnote{\kern 0.5pt \raise 0.5ex \hbox{$\dag$}}{%
In the ``Redux'' section we examine more closely what ``knowing the value on every $x$'' means.}

From equation $(**)$ we see that the value of the function $\E{f \mid {\cal F}}$ 
evaluated on any `x' value in a minimal set, $\Lambda$, is% 
\footnote{\kern 0.5pt \raise 0.5ex \hbox{\ddag}}{%
The notation, $P_{\Lambda}$ means that we have normalized the measure, $P$, 
so that $P_{\Lambda}(\Lambda) = 1$. To do this we are {\it assuming\/} 
that $P(\Lambda) \ne 0$.}
$$
\eqalignno{
	\E{f \mid {\cal F}}\!(x) & = \frac{\int_\Lambda f \, dP}{P(\Lambda)} = \int_\Lambda f \, dP_{\Lambda} & (4) \cr
}
$$
That is, the value of the conditional expectation on any minimal set is the 
{\it weighted average\/} of $f$ over the minimal set. The weights are 
determined by normalizing the probability measure $P$ over the minimal set.

This is the view that was promised at the beginning of this paper.


\section{Example Redux}
Let's adjust the die example probabilities and redo the calculation of the
conditional expectation. Replace the uniform $P$ from before with the new
probability measure:
$$
\eqalignno{
	P(A) & = \frac{|A\cap \{D_1,D_3,D_5\}^c|}{3} \quad A \in {\cal E} & \cr
}
$$
That is, the probability of a set, $A$, is the number of dice in $A$ that
are not in the set $\{D_1, D_3, D_5\}$ divided by $3$.
With this definition, $P(\Lambda_1) = 0$ and $P(\Lambda_2) = 1$.

Given $P$, we can find the value of $\E{f \mid {\cal F}}$ on the set $\Lambda_2$ 
as before using equation (4).
However, we can't use it to find the value $\E{f \mid {\cal F}}$ on the 
set $\Lambda_1$ as $P(\Lambda_1) = 0$. 

It turns out, we don't have to determine the values with any specificity on $\Lambda_1$.
We can {\it set\/} the values of $\E{f \mid {\cal F}}$ on $\Lambda_1$ to $0$; or, 
to any other {\it single\/} value for that matter%
\footnote{\kern 0.5pt \raise 0.5ex \hbox{\dag}}{%
We know that $\Lambda_1$ is minimal; so, for the conditional expectation 
function to be measurable it must have the same value on all elements of $\Lambda_1$.}.
The reason for this is 
that the definition of conditional expectation determines a unique element in 
the function space on ${\cal F}$.
However, each element in such a space is actually an equivalence class of 
measurable functions over ${\cal F}$ that differ on a set of measure $0$. 

We can't directly compute what the value of a representative of $\E{f \mid {\cal F}}$ is on 
the set $\Lambda_1$, but we don't have to, since it's a set of measure $0$, any 
value on this set will give us a function that is in the equivalence class of 
the conditional expectation element.
Therefore, we may take $\E{f \mid {\cal F}}$ to be $0$ on $\Lambda_1$. This gives us 
a representative for the conditional expectation as we have values for all $x \in X$.
Through this representative we can produce the associated equivalence class of functions.
Consequently, we know $\E{f \mid {\cal F}}$ in the function space $L^1(X, {\cal F}, P)$.

In the case of singular sets (a set of probability $0$) we may revise equation (4) and write
that on minimal sets, $\E{f \mid {\cal F}}$ is constant and satisfies:

$$
\eqalignno{
	\E{f \mid {\cal F}}\!(x) \cdot P(\Lambda) & = \int_\Lambda f \, dP & (5) \cr
}
$$
Note that when $P(\Lambda) = 0$ both sides of (5) vanish identically, so (5) imposes
no constraint on the value of $\E{f \mid {\cal F}}$ on $\Lambda$ -- consistent with the
freedom (afforded by the equivalence-class viewpoint) to assign that value arbitrarily.

Here, with a measure theoretic methodology, we have an elegant way to talk about 
conditional expectation in a singular context that doesn't exist in the 
traditional approach.

\appendix{A Discrete Conditional Expectation Result}
We generalize the discussion from the previous sections and work with functions
on discrete sets with a given algebra of sets.

{\bf Definition:} Given a discrete set $X$ and an algebra of sets ${\cal F}$,
a set $A$ is minimal in $X$ with respect to ${\cal F}$ iff $A \in {\cal F}$
and no non-empty proper subset of $A$ is contained in ${\cal F}$.

Let $X$ be a set with a finite number of elements.
For any given $\sigma$-algebra, ${\cal F}$, (in this case algebra) of sets of $X$, 
let $M_{\cal F} = \{M_i\}_{i=1}^N$
be the set of all {\it distinct\/} minimal sets of $X$ in ${\cal F}$. 
The set $M_{\cal F}$ 
exists as we are dealing with a finite set $X$ with a finite power set $2^X$. 
The set $M_{\cal F}$ is formed simply by examining each element in the power 
set of $X$ and adding it to $M_{\cal F}$ if it is minimal in ${\cal F}$.

{\bf Claim:\/} 
The set $M_{\cal F}$ is a {\it partition\/} of the set $X$.
Meaning:

\beginEnum
	\item{$\forall i, j \in \{1, \ldots, N\} \quad M_i \cap M_j = \emptyset \quad (i \ne j)$;}
	\item{$X = \mathop{\cup}_{i=1}^N\limits M_i$.}
	\item{The algebra generated by the sets in $M_{\cal F}$ is the collection 
			of the unions of each 
			element in the power set of $M_{\cal F}$.
		Further, the algebra generated is ${\cal F}$.}
\endEnum
We prove the first property by contradiction. Suppose $M_i \ne M_j$ are both
minimal in ${\cal F}$ with $M_i \cap M_j \ne \emptyset$. Since ${\cal F}$ is an
algebra, $M_i \cap M_j \in {\cal F}$. It is a non-empty subset of $M_i$, so by
minimality of $M_i$ it must equal $M_i$; hence $M_i \subseteq M_j$. By
minimality of $M_j$, this in turn gives $M_i = M_j$, contradicting $M_i \ne M_j$.

We prove the second property by contradiction. If the union of minimal sets did not span $X$,
then the set, $Z_0 = \bigl\{\mathop{\cup}_{i=1}^N\limits M_i\bigr\}^c$, is non-empty. 
The set $Z_0$ cannot be minimal as we have already accounted for all minimal 
sets in our collection. Set ${\cal H} = 2^{Z_0} \setminus \{\emptyset, Z_0\}$ -- the
power set of $Z_0$ less $Z_0$ and $\emptyset$. Note that for $Z_0$ to be non-minimal, it 
cannot consist of $1$ element. Therefore, ${\cal H}$ is a non-empty collection 
of sets. If none of the non-empty sets are in ${\cal F}$, then $Z_0$ is 
minimal  -- a contradiction. Therefore, there must be a non-trivial strict subset of 
$Z_0$, $Z_1$, that is in ${\cal F}$. We can repeat this argument with $Z_1$ to 
produce, $Z_2$, $\ldots Z_i$ producing $Z_{i+1}$, etc; noting, that each time this
is done the number of elements in the current ``$Z$ set'' decreases by at least $1$. 
Consequently, the process cannot proceed indefinitely. Let $m$ be the step 
where this process fails; that is, there is no strict subset, $Z_m$, of set $Z_{m-1}$
which is a subset of ${\cal F}$. But this means that 
$Z_m$ is minimal by definition -- contradiction.
Consequently, the second property holds.

The third property follows as none of the $\mathop{\{M_i\}}_{i=1}^N$ can produce sets of 
finer granularity than themselves; they behave as if they were single elements. Therefore,
the only way to produce new sets is by union, and consequently, the collection of 
all elements generated is the collection of the union of each element from the
power set of $\mathop{\{M_i\}}_{i=1}^N$.
Finally, if we label ${\cal G}$ as the algebra generated by $M_{\cal F}$, then since
its generator sets, $\{M_i\}_{i=1}^N$ are elements of ${\cal F}$, it must be the
case that ${\cal G} \subseteq {\cal F}$. To show equality, suppose
that there is a set $F \in {\cal F}$ but $F \notin {\cal G}$. But
$F = F \cap X = F \cap \mathop{\cup}_{i=1}^N\limits M_i = \mathop{\cup}_{i=1}^N\limits (F \cap M_i)$.
Since for a given $i$, $M_i$ is minimal in ${\cal F}$, and $F \in {\cal F}$,
the intersection must be either $M_i$ or $\emptyset$. Consequently, $F$ is
a union of an element of the power set of $\{M_i\}_{i=1}^N$, which has
been shown to be a member of ${\cal G}$. Consequently, ${\cal G} = {\cal F}$.

Let $(X, {\cal E}, P)$ be a probability space with $X$ discrete, let $f$ be measurable on it,
and let ${\cal F}$ be a sub $\sigma$-algebra of ${\cal E}$. We
examine the expectation of the conditional expectation, $\E{f / {\cal F}}$.%
\footnote{\kern 0.5pt \raise 0.5ex \hbox{\dag}}{%
Remember, the conditional expectation is not a number; it is a random variable
(a measurable function on $(X, {\cal F})$). As such, it too has an expectation.
}

$$
\eqalignno{
E[\E{f / {\cal F}}] & = \int_{X} \E{f / {\cal F}} \, dP_{\cal F}  = \int_{X} f \, dP & (6) \cr
     & = \int_{\mathop{\cup}_{i=1}^N\limits M_i} f \, dP & \cr
     & = \sum_{i=1}^N \int_{M_i} f \, dP & \cr
     & = \sum_{i=1}^N \int_{M_i} \E{f / {\cal F}} \, dP_{\cal F} &\cr
     & = \sum_{i=1}^N \E{f / {\cal F}}\!(M_i)\, P(M_i) & (7) \cr
}
$$
In equation (6) we have the following identity: $E[\E{f /{\cal F}}]= E[f]$.
From equation (7) we have that the expectation of the conditional expectation
of a $\sigma$-algebra generated by minimal sets is just the weighted sum of
the values of the conditional expectation on these sets.

Finally, if $g$ is a discrete function, the collection of all non-empty sets
of the form $g^{-1}(a)$, $a \in (-\infty, \infty)$,
is a finite collection of sets, $\mathop{\{M_i\}}_{i=1}^N$, for some positive
integer $N$.%
\footnote{\kern 0.5pt \raise 0.5ex \hbox{\ddag}}{%
	The $N$ sets correspond to $N$ distinct values
	$\mathop{\{a_i\}}_{i=1}^N$, with each $M_i = g^{-1}(a_i)$ and each $a_i$ a real number.
}
The collection of the union of
each element from the power set of these sets
generates an algebra, ${\cal F}$, with each $M_i\; (1 \le i \le N)$ being
the minimal sets in this algebra. We write $\E{f \mid g}$ to mean $\E{f / {\cal F}}$.
Consequently, we may write the above as:

$$
\eqalignno{
	E[\E{f \mid g}] & = \int_{X} \E{f \mid g} \, dP = \int_{X} f \, dP & (6') \cr
									         & = \int_{\mathop{\cup}_{i=1}^N\limits M_i} f \, dP & \cr
											 & = \sum_{i=1}^N \int_{M_i} f \, dP & \cr
											 & = \sum_{i=1}^N \int_{M_i} \E{f \mid g} \, dP &\cr
											 & = \sum_{i=1}^N \E{f \mid g}(M_i)\, P(M_i) & (7') \cr
}
$$
Here, the term, $\E{f \mid g}(M_i)$ is interpreted as the value of $\E{f / {\cal F}}$ on {\it any\/} element of the set $M_i$ -- 
as the value of the function $\E{f / {\cal F}}$ is constant on $M_i$.
We also drop the $dP_{\cal F}$ notation, which was only used as a crutch in the previous part of the paper.
Lastly, we know from equation (5) how to compute the values of $\E{f \mid g}$ for all $M_i$ -- even when the probability of
a ``g'' event is $0$.

Equation (6') says that $E[\E{f \mid g}] = E[f]$; that is, the expectation of the conditional expectation of $f$
is the same as the expectation of $f$. Equation (7') states that on each level set of $g$,
$M_i = g^{-1}(a_i)$, the conditional expectation $\E{f \mid g}$ is the $P$-weighted average of $f$ over $M_i$;
summing those values weighted by $P(M_i)$ recovers $E[f]$.


\bye

