\documentclass{article}

\usepackage{amsthm, amsmath, amssymb}
\usepackage{listings}
\usepackage{xcolor}

% VS Code "Default Light+" inspired colors for Lean 4.
\definecolor{vscBg}{RGB}{255,255,255}        % white background
\definecolor{vscText}{RGB}{0,16,128}         % default identifier (dark blue)
\definecolor{vscKeyword}{RGB}{0,0,255}       % blue: def, theorem, axiom, ...
\definecolor{vscControl}{RGB}{175,0,219}     % purple: by, intro, exact, ...
\definecolor{vscComment}{RGB}{0,128,0}       % green: comments
\definecolor{vscString}{RGB}{163,21,21}      % dark red: strings
\definecolor{vscType}{RGB}{38,127,153}       % teal: Nat, Set, Prop, Finset, ...
\definecolor{vscFunction}{RGB}{121,94,38}    % brown: function names
\definecolor{vscNumber}{RGB}{9,134,88}       % green-blue: numerals
\definecolor{vscOperator}{RGB}{0,0,0}        % black: operators

\lstdefinelanguage{Lean4}{
  % Group 1: declaration keywords (blue)
  morekeywords=[1]{def,theorem,lemma,axiom,abbrev,instance,class,structure,
    inductive,namespace,end,open,import,attribute,deriving,where,variable,
    section},
  % Group 2: tactic and control keywords (purple)
  morekeywords=[2]{by,intro,intros,exact,apply,refine,rcases,obtain,simp,ext,
    sorry,rfl,decide,omega,ring,norm_num,dsimp,rw,set,constructor,rintro,
    interval_cases,simp_all,have,show,from,fun,let,in,if,then,else,match,
    with,this,calc,assumption,push_neg,by_cases,classical,split,refl,
    induction,cases},
  % Group 3: type names and important identifiers (teal)
  morekeywords=[3]{Nat,Int,Set,Prop,Type,Sort,Finset,Iff,And,Or,Not,Exists,
    Membership,Mem,Function,Subtype,Option,List,Array,Bool,True,False,
    Pythagorean,Prime,Infinite,IsHyp,Hd,Hp,Hu,H,PrimPythHyp,Finsupp},
  morecomment=[s][\color{vscComment}\itshape]{/-}{-/},
  morecomment=[l][\color{vscComment}\itshape]{--},
  morestring=[b][\color{vscString}]",
  basicstyle=\ttfamily\footnotesize\color{vscText},
  keywordstyle=[1]\color{vscKeyword},
  keywordstyle=[2]\color{vscControl},
  keywordstyle=[3]\color{vscType},
  identifierstyle=\color{vscText},
  showstringspaces=false,
  columns=fullflexible,
  keepspaces=true,
  literate=
    {ℕ}{{{\color{vscType}$\mathbb{N}$}}}1
    {ℤ}{{{\color{vscType}$\mathbb{Z}$}}}1
    {𝓗}{{$\mathcal{H}$}}1
    {∀}{{$\forall$}}1
    {∃}{{$\exists$}}1
    {∈}{{$\in$}}1
    {∉}{{$\notin$}}1
    {∧}{{$\wedge$}}1
    {∨}{{$\vee$}}1
    {¬}{{$\neg$}}1
    {→}{{$\to$}}1
    {←}{{$\leftarrow$}}1
    {↔}{{$\leftrightarrow$}}1
    {∏}{{$\prod$}}1
    {⇒}{{$\Rightarrow$}}1
    {≠}{{$\neq$}}1
    {≤}{{$\leq$}}1
    {≥}{{$\geq$}}1
    {₀}{{$_0$}}1
    {₁}{{$_1$}}1
    {₂}{{$_2$}}1
    {²}{{$^2$}}1
    {⟨}{{$\langle$}}1
    {⟩}{{$\rangle$}}1
    {↦}{{$\mapsto$}}1
    {·}{{$\cdot$}}1
    {×}{{$\times$}}1
    {ˢ}{{$^{\textrm{s}}$}}1
    {≡}{{$\equiv$}}1
    {∪}{{$\cup$}}1
    {∩}{{$\cap$}}1
    {⊆}{{$\subseteq$}}1
    {∅}{{$\emptyset$}}1
    {∣}{{$\mid$}}1
}
\lstnewenvironment{leancode}{\lstset{language=Lean4}}{}

\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.0in}
\setlength{\hoffset}{-0.75in}
\setlength{\voffset}{-0.75in}

\setlength{\parindent}{0.0in}
\setlength{\parskip}{1.5\baselineskip}

\title{Conjectures for Primitive Pythagorean Triples}
\author{R. Scott McIntire}
\date{May 1, 2026}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}

\theoremstyle{definition}
\newtheorem{defn}{Definition}


\begin{document}
\maketitle


\section{Overview}
We discuss the set ${\cal H}$ of primitive Pythagorean hypotenuses --
the ``hypotenuse'' values that arise from primitive Pythagorean triples
(right triangles with integer sides sharing no common factor) --
recall the classical theorems describing its structure, and state
an open conjecture about the cumulative parity of its prime elements.

A related set, ${\cal H}_p$, is the subset of ${\cal H}$ that are primes. 
That is, primitive Pythagorean hypotenuses that are prime.

Two of the three observations below turn out to be classical theorems —
we recall them for orientation — while the third remains an open conjecture
of mine, which is the focus of this note.
\begin{enumerate}
    \item \textbf{Partition of ${\cal H}$} (theorem): ${\cal H}$ is {\em partitioned\/}
      into two disjoint infinite sets: the set of all powers of Pythagorean primes,
      and the set of all hypotenuses arising from more than one primitive Pythagorean triple.
    \item \textbf{Infinitude of ${\cal H}_p$} (Dirichlet): The set of Pythagorean
      primes is infinite.
    \item \textbf{${\cal H}_p \sim \mathbb{P}$} (conjecture): ${\cal H}_p$ is {\em similar\/}
      to and as {\em complex\/} as the primes when measured by certain
      cumulative-parity metrics.
\end{enumerate}


\section{Definitions}
After a few definitions we describe a conjecture regarding Primitive 
Pythagorean Triples. We use standard notation, associating the natural 
numbers with the symbol $\mathbb{N}$%
\footnote{We assume that $\mathbb{N}$ does {\em not\/} include $0$.}
and we use the symbol $\mathbb{P}$ to denote the set of prime numbers.


\begin{defn}
{\em Pythagorean Triples\/}, denoted ${\cal T}$, are triples of
numbers representing the lengths of the sides of right triangles where all the sides are integers.%
\footnote{Triples are taken as ordered, so $(3,4,5)$ and $(4,3,5)$ are
distinct elements of ${\cal T}$.}
Specifically,
\begin{eqnarray}
  {\cal T} & = & \{ (x,y,z) \; | \; (x,y,z) \in \mathbb{N}^3 \; \wedge \; x^2 + y^2 = z^2 \}.
\end{eqnarray}
\end{defn}

\begin{defn}
{\em Primitive Pythagorean Triples\/}, denoted ${\cal T}_p$, are
Pythagorean triples whose legs share no common factor:%
\footnote{Since any common divisor of $x$ and $y$ also divides $z$,
$\gcd(x,y) = 1$ is equivalent to $\gcd(x,y,z) = 1$, the usual primality condition.}
\begin{eqnarray}
  {\cal T}_p & = & \{ (x,y,z) \; | \; (x,y,z) \in \mathbb{N}^3 \; \wedge \; \gcd(x,y) = 1 \; \wedge \; x^2 + y^2 = z^2 \}.
\end{eqnarray}
\end{defn}

\begin{defn}
{\em Primitive Pythagorean Hypotenuses\/}, denoted ${\cal H}$, are defined by:
\begin{eqnarray}
  {\cal H} & = & \{ z \; | \; \exists (x,y) \in \mathbb{N}^2, \;  (x,y,z) \in {\cal T}_p \}.
\end{eqnarray}
This is the set of all possible hypotenuses of primitive Pythagorean triples.
\end{defn}

\begin{defn}
{\em Pythagorean Primes\/}, denoted ${\cal H}_p$, are defined by:
\begin{eqnarray}
  {\cal H}_p & = & \{ z \; | \; \exists (x,y) \in \mathbb{N}^2, \, (x,y,z) \in {\cal T}_p \; \wedge \; z \in \mathbb{P} \}.
\end{eqnarray}
This is the set of primes that occur as hypotenuses of primitive Pythagorean triples.
\end{defn}

\begin{defn}
{\em Duplicate Primitive Hypotenuses\/}, denoted ${\cal H}_d$, are defined by:
\begin{eqnarray}
  {\cal H}_{d} & = & \{ z \; | \; \exists\, (x_1, x_2, y_1, y_2) \in \mathbb{N}^4, \;
      (x_1,y_1,z) \in {\cal T}_p \; \wedge \; (x_2, y_2, z) \in {\cal T}_p \; \wedge \; \{x_1, y_1\} \neq \{x_2, y_2\} \}.
\end{eqnarray}
This is the set of primitive hypotenuses that arise from more than one
{\em distinct\/} primitive Pythagorean triple (i.e.\ not merely a swap of legs).
\end{defn}

\begin{defn}
{\em Primitive Hypotenuses Powers\/}, denoted ${\cal H}_u$, are defined by:
\begin{eqnarray}
  {\cal H}_u & = & \{ z \; | \; \exists\,(p,n) \in {\cal H}_p \times \mathbb{N}, \; z = p^n \}.
\end{eqnarray}
This is the set of all powers of Pythagorean primes.
\end{defn}


\begin{defn}
The {\em Prime Parity\/} function, $\Pi$, is defined on any prime $p$ with $p \ge 5$ by:
\begin{eqnarray}
  \Pi(p) = \begin{cases}
  \hphantom{-}1\hfil & \exists n \in \mathbb{N}, p = 6n + 1 \\
            -1 \hfil & \exists n \in \mathbb{N}, p = 6n - 1 \\
             \end{cases}
\end{eqnarray}
\end{defn}

\begin{defn}
The {\em First Deviation\/} function, $\Delta_p$, is defined by:%
\footnote{If for a given $n$,
$\left\{ m \; \left| \; n \le \left( \sum_{i=1}^m \Pi({\cal H}_p[i]) - \sum_{i=3}^{m+2} \Pi(\mathbb{P}[i]) \right. \right) \right\} = \emptyset$, we set $\Delta_p(n) = \infty$.}
\begin{eqnarray}
\Delta_p(n) = \min_{m} \left\{ m \; \left| \; n \le \left( \sum_{i=1}^m \Pi({\cal H}_p[i]) - \sum_{i=3}^{m+2} \Pi(\mathbb{P}[i])\right. \right) \right\}
\end{eqnarray}
\end{defn}

\newpage
\textbf{Notes:}
\begin{itemize}
  \item The subscript $u$ in the name ${\cal H}_u$ is not yet justified.
  \item ${\cal H}_p$ is a {\em proper\/} subset of the primes: ${\cal H}_p \subset \mathbb{P}$.
  \item If the sets above are treated as lists, they are indexed in sorted order.
  \item Clearly, $\Delta_p$ is a non-decreasing function.
\end{itemize}

First we list a few facts and state a couple of known theorems before proceeding:
\begin{enumerate}
  \item The Pythagorean primes are of the form $4n + 1$.
      This is a special case of a more general theorem proven by
        Pierre de Fermat, known as Fermat's theorem on sums of two squares.
      According to this theorem, an odd prime number $p$ can be expressed as
        the sum of two squares if and only if
      $p$ is congruent to $1$ modulo $4$, i.e., $p = 4n + 1$ for some
        positive integer $n$.
  \item Moreover, the distribution of primes of the form $4n + 1$ has
      been studied extensively.
      The German mathematician Peter Gustav Lejeune Dirichlet proved that
        there are infinitely many primes of the form $4n + 1$, a result known as
      Dirichlet's theorem on arithmetic progressions.
      This theorem, which generalizes the case of Pythagorean primes, states
        that for any two positive coprime numbers $a$ and $d$,
      there are infinitely many primes of the form $a + nd$, where $n$ is a
        non-negative integer.
  \item In the case of Pythagorean primes, we can take $a = 1$ and $d = 4$ to
      get the form $4n + 1$.
        Dirichlet's theorem thus guarantees that there are infinitely
        many Pythagorean primes.
      This is a profound result, as it extends the prime number theorem,
      which states that there are infinitely many prime numbers, to
      arithmetic progressions.
\end{enumerate}
From these results we get the following characterization of Pythagorean primes:
\begin{theorem}
${\cal H}_p = \{ p \; | \; p \in \mathbb{P} \; \wedge \; p \equiv 1 \pmod{4}\}$.
\end{theorem}
And the following structure theorem for Pythagorean hypotenuses:
\begin{theorem}
The set ${\cal H}$ has a unique factorization (up to ordering) in ${\cal H}_p$:
each element of ${\cal H}$ can be uniquely written as a product of powers from ${\cal H}_p$,
\begin{eqnarray}
  \forall h \in {\cal H},\, \exists ! N \in \mathbb{N}, \, \exists ! ({\bf p},{\bf k}) \in \left({\cal H}_p^{N} \times \mathbb{N}^{N}\right): h = \prod_{i=1}^N p_{i}^{k_i},
\end{eqnarray}
where uniqueness is up to permutation of the index $i$.
\end{theorem}

A classical consequence is the following partition theorem, which I had originally
stated as a conjecture but which follows from the count of primitive
representations of $n$ as a sum of two squares:
\begin{theorem}\label{thm:partition}
The set ${\cal H}$ is partitioned by ${\cal H}_d$ and ${\cal H}_u$:
\begin{enumerate}
  \item $\hphantom{|{\cal H}_d|}{\cal H} = {\cal H}_{d} \cup {\cal H}_u$;
  \item $\hphantom{|{\cal H}_d|}\emptyset = {\cal H}_{d} \cap {\cal H}_u$;
  \item $\hphantom{\emptyset}|{\cal H}_d| = \infty$.
\end{enumerate}
\end{theorem}

\textbf{Proof sketch.} For $n \in {\cal H}$, the number of primitive Pythagorean
triples with hypotenuse $n$ equals $2^{k-1}$, where $k$ is the number of
distinct prime factors of $n$ (all of which lie in ${\cal H}_p$ by Theorem~2).
This count is a classical consequence of the unique factorization of $n$ in
the Gaussian integers. Hence $n$ has exactly one primitive representation iff
$k = 1$, i.e.\ iff $n$ is a power of a single Pythagorean prime ($n \in {\cal H}_u$);
otherwise $n \in {\cal H}_d$. This gives the partition. The infinitude of
${\cal H}_d$ follows from the infinitely many distinct products $p \cdot q$
with $p, q \in {\cal H}_p$ and $p \neq q$ (Dirichlet).\hfill$\square$

\newpage
\section{McIntire's Pythagorean Triple Conjectures}

The remaining open conjecture concerns the {\em cumulative parity\/} of
${\cal H}_p$ — the running sum $\sum_{i=1}^n \Pi({\cal H}_p[i])$ — and how
it compares to the corresponding sum over all primes $\geq 5$.

\begin{conjecture}\label{conj:parity}
The cumulative parity of ${\cal H}_p$ exhibits the following non-trivial behavior:
\begin{enumerate}
  \item $\lim\limits_{n \rightarrow \infty} \dfrac{\sum\limits_{i=1}^n \Pi({\cal H}_p[i])}{n} = 0$;
  \item $\forall n\in \mathbb{N}, \Delta_p(n) < \infty$;
  \item $\lim\limits_{n \rightarrow \infty} \dfrac{\Delta_p(n)}{n^{\ln(2 \pi)}} = 41$.%
\footnote{The exponent $\ln(2\pi) \approx 1.838$ and the constant $41$ are
based on numerical fitting; we do not have a heuristic derivation. Readers
should treat the precise values as empirical, subject to revision as more
data becomes available.}
\end{enumerate}
\end{conjecture}

The three sub-conjectures together suggest that the Pythagorean primes
are ``as complex as'' the full set of primes when measured by their
parity behavior. Sub-conjecture~1 says the $\pm 1$ values $\Pi({\cal H}_p[i])$
average to zero, so ${\cal H}_p$ is spread evenly between residues $1$ and
$-1 \pmod 6$ in the long run, mirroring $\mathbb{P}$. Numerically, however,
the cumulative parity of ${\cal H}_p$ tends to run lower than that of
$\mathbb{P}$; sub-conjectures~2 and~3 quantify how often and how far the
${\cal H}_p$ sum nonetheless overtakes the $\mathbb{P}$ sum, asserting that
deviations of any size $n$ do occur ($\Delta_p(n) < \infty$) and that they
grow at the polynomial rate $n^{\ln(2\pi)}$. This ``mixing'' is what we mean
by saying ${\cal H}_p$ is non-trivial in a way similar to $\mathbb{P}$.

\newpage
\appendix
\section{A formally verified Lean 4 proof of Theorem 3}

The proof below has been mechanically checked by Lean 4
(toolchain \verb|leanprover/lean4:v4.29.1|) against Mathlib.
The full project lives at \verb|lean/PythHyp/| in this repository; the proof
file is\\\verb|lean/PythHyp/PythHyp/Basic.lean|.

The decomposition is honest: three classical results — explicitly named
(F), (G-easy), and (G-uniq) — are stated as \verb|axiom|s rather than
proved in this file, because their formalization, while certainly
possible in Mathlib, is a substantial project of its own. They are
classical theorems, not conjectures:
\begin{itemize}
  \item[(F)] {\em Fermat's characterization.\/} A positive integer $z > 1$
    is a primitive Pythagorean hypotenuse iff every prime factor of $z$ is
    $\equiv 1 \pmod 4$. The prime case is in Mathlib as
    \verb|Nat.Prime.sq_add_sq|; the general case follows from
    multiplicativity of the norm in $\mathbb Z[i]$.
  \item[(G-easy)] {\em Multiple factors yield duplicates.\/} If $z \in
    \mathcal H$ has at least two distinct prime factors, then $z$ admits
    at least two distinct unordered primitive representations
    $a^2 + b^2 = z^2$. This is the easy half of the classical
    $2^{k-1}$ count and follows from a single non-trivial
    factorization of $z$ in $\mathbb Z[i]$.
  \item[(G-uniq)] {\em Prime powers are unique.\/} If $z = p^n$ for a
    Pythagorean prime $p$ and $n \ge 1$, then $z$ admits only one
    unordered primitive representation. This is the harder half of the
    Gauss count.
\end{itemize}
Dirichlet's theorem on primes $\equiv 1 \pmod 4$ is {\em not\/} an axiom
in this proof: it is invoked directly as the Mathlib lemma
\verb|Nat.infinite_setOf_prime_modEq_one|.

Modulo the three classical inputs (F), (G-easy), (G-uniq), every step
of Theorem 3 — parts (a), (b), and (c) — is fully formalized, with no
\verb|sorry|. Lean reports a clean build.

\begin{leancode}
import Mathlib.NumberTheory.PythagoreanTriples
import Mathlib.NumberTheory.SumTwoSquares
import Mathlib.NumberTheory.Zsqrtd.GaussianInt
import Mathlib.NumberTheory.PrimesCongruentOne
import Mathlib.Data.Nat.PrimeFin

open Set

namespace PrimPythHyp

/-- z is a primitive Pythagorean hypotenuse. -/
def IsHyp (z : ℕ) : Prop :=
  ∃ a b : ℕ, 0 < a ∧ 0 < b ∧ Nat.gcd a b = 1 ∧ a^2 + b^2 = z^2

def H : Set ℕ := { z | IsHyp z }
def Hp : Set ℕ := { z | z ∈ H ∧ z.Prime }

/-- Hypotenuses with at least two distinct unordered primitive
    representations. -/
def Hd : Set ℕ :=
  { z | ∃ a₁ b₁ a₂ b₂ : ℕ,
      0 < a₁ ∧ 0 < b₁ ∧ Nat.gcd a₁ b₁ = 1 ∧ a₁^2 + b₁^2 = z^2 ∧
      0 < a₂ ∧ 0 < b₂ ∧ Nat.gcd a₂ b₂ = 1 ∧ a₂^2 + b₂^2 = z^2 ∧
      ({a₁, b₁} : Finset ℕ) ≠ {a₂, b₂} }

/-- Powers of Pythagorean primes. -/
def Hu : Set ℕ := { z | ∃ p ∈ Hp, ∃ n : ℕ, 0 < n ∧ z = p^n }

----------------------------------------------------------------------
-- Classical building blocks (axiomatized; these are real theorems)
----------------------------------------------------------------------

/-- (F) Fermat's two-squares + multiplicativity in ℤ[i]. -/
axiom mem_H_iff_factors {z : ℕ} (hz : 1 < z) :
    z ∈ H ↔ ∀ p ∈ z.primeFactors, p % 4 = 1

/-- (G-easy) ≥ 2 distinct prime factors ⇒ ≥ 2 distinct primitive reps. -/
axiom multiple_factors_yields_Hd {z : ℕ} (hz : z ∈ H)
    (hk : 2 ≤ z.primeFactors.card) : z ∈ Hd

/-- (G-uniq) prime power ⇒ only one primitive rep. -/
axiom prime_power_not_in_Hd {p : ℕ} (hp : p ∈ Hp) {n : ℕ} (hn : 0 < n) :
    p^n ∉ Hd

----------------------------------------------------------------------
-- Helper
----------------------------------------------------------------------

lemma one_lt_of_in_H {z : ℕ} (hz : z ∈ H) : 1 < z := by
  obtain ⟨a, b, ha, hb, _, hab⟩ := hz
  have h1a : 1 ≤ a^2 := Nat.one_le_iff_ne_zero.mpr (pow_ne_zero 2 ha.ne')
  have h1b : 1 ≤ b^2 := Nat.one_le_iff_ne_zero.mpr (pow_ne_zero 2 hb.ne')
  have h2 : 2 ≤ z^2 := by rw [← hab]; omega
  rcases Nat.lt_or_ge z 2 with hlt | hge
  · interval_cases z <;> simp_all
  · omega

----------------------------------------------------------------------
-- Theorem 3
----------------------------------------------------------------------

/-- (a) H = Hd ∪ Hu. -/
theorem H_eq_Hd_union_Hu : H = Hd ∪ Hu := by
  ext z
  refine ⟨fun hz => ?_, ?_⟩
  · have hz1 : 1 < z := one_lt_of_in_H hz
    have hz_ne : z ≠ 0 := by omega
    have hcard_pos : 1 ≤ z.primeFactors.card := by
      have : z.primeFactors.Nonempty := by
        rw [Nat.nonempty_primeFactors]; exact hz1
      exact Finset.card_pos.mpr this
    by_cases hk : z.primeFactors.card = 1
    · -- One prime factor: z = p^n.
      right
      obtain ⟨p, hp_eq⟩ := Finset.card_eq_one.mp hk
      have hp_in : p ∈ z.primeFactors := by
        rw [hp_eq]; exact Finset.mem_singleton_self p
      have hp_prime : p.Prime := Nat.prime_of_mem_primeFactors hp_in
      have hp_dvd : p ∣ z := Nat.dvd_of_mem_primeFactors hp_in
      set n := z.factorization p
      have hn_pos : 0 < n :=
        hp_prime.factorization_pos_of_dvd hz_ne hp_dvd
      have hz_eq : z = p ^ n := by
        have hprod : z.factorization.prod (· ^ ·) = z :=
          Nat.prod_factorization_pow_eq_self hz_ne
        have hsupp : z.factorization.support = {p} := by
          rw [Nat.support_factorization, hp_eq]
        rw [← hprod, Finsupp.prod, hsupp, Finset.prod_singleton]
      refine ⟨p, ⟨?_, hp_prime⟩, n, hn_pos, hz_eq⟩
      rw [mem_H_iff_factors hp_prime.one_lt]
      intro q hq
      rw [hp_prime.primeFactors, Finset.mem_singleton] at hq
      rw [hq]
      exact (mem_H_iff_factors hz1).mp hz p hp_in
    · -- ≥ 2 prime factors.
      left
      have hk2 : 2 ≤ z.primeFactors.card := by omega
      exact multiple_factors_yields_Hd hz hk2
  · rintro (hd | hu)
    · obtain ⟨a, b, _, _, ha, hb, hgcd, heq, _⟩ := hd
      exact ⟨a, b, ha, hb, hgcd, heq⟩
    · obtain ⟨p, hpH, n, hn, rfl⟩ := hu
      have hp_prime : p.Prime := hpH.2
      have hpn_gt1 : 1 < p^n := one_lt_pow₀ hp_prime.one_lt hn.ne'
      rw [mem_H_iff_factors hpn_gt1]
      intro q hq
      have hpf : (p^n).primeFactors = {p} :=
        Nat.primeFactors_prime_pow hn.ne' hp_prime
      rw [hpf, Finset.mem_singleton] at hq
      rw [hq]
      exact (mem_H_iff_factors hp_prime.one_lt).mp hpH.1 p
        hp_prime.mem_primeFactors_self

/-- (b) Hd ∩ Hu = ∅. -/
theorem Hd_inter_Hu : Hd ∩ Hu = ∅ := by
  ext z
  simp only [mem_inter_iff, mem_empty_iff_false, iff_false, not_and]
  rintro hd ⟨p, hpH, n, hn, rfl⟩
  exact prime_power_not_in_Hd hpH hn hd

/-- (c) Hd is infinite. -/
theorem Hd_infinite : Hd.Infinite := by
  set S : Set ℕ := {p | p.Prime ∧ p ≡ 1 [MOD 4]} \ {5}
  have h_pyth : Set.Infinite {p : ℕ | p.Prime ∧ p ≡ 1 [MOD 4]} :=
    Nat.infinite_setOf_prime_modEq_one (by norm_num)
  have hS_inf : S.Infinite := h_pyth.diff (Set.finite_singleton _)
  apply Set.infinite_of_injOn_mapsTo
    (s := S) (t := Hd) (f := fun p => 5 * p)
  · intro p₁ _ p₂ _ h; dsimp only at h; omega
  · intro p hpS
    obtain ⟨⟨hp_prime, hp_mod⟩, hp_ne⟩ := hpS
    have hp_ne_5 : p ≠ 5 := fun h => hp_ne (by simp [h])
    have h5_prime : Nat.Prime 5 := by decide
    have hp_pos : 0 < p := hp_prime.pos
    have hp_ne0 : p ≠ 0 := hp_pos.ne'
    have h5p_gt1 : 1 < 5 * p := by
      have : 1 < p := hp_prime.one_lt; omega
    have h5p_in_H : 5 * p ∈ H := by
      rw [mem_H_iff_factors h5p_gt1,
          Nat.primeFactors_mul (by norm_num) hp_ne0,
          h5_prime.primeFactors, hp_prime.primeFactors]
      intro q hq
      rw [Finset.mem_union, Finset.mem_singleton,
          Finset.mem_singleton] at hq
      rcases hq with rfl | rfl
      · decide
      · exact hp_mod
    have hcard : 2 ≤ (5*p).primeFactors.card := by
      have h5_in : (5 : ℕ) ∈ (5*p).primeFactors :=
        Nat.mem_primeFactors.mpr ⟨h5_prime, ⟨p, rfl⟩, by omega⟩
      have hp_in : p ∈ (5*p).primeFactors :=
        Nat.mem_primeFactors.mpr ⟨hp_prime, ⟨5, by ring⟩, by omega⟩
      have hsub : ({5, p} : Finset ℕ) ⊆ (5*p).primeFactors := by
        intro x hx
        rw [Finset.mem_insert, Finset.mem_singleton] at hx
        rcases hx with rfl | rfl <;> assumption
      have h2 : ({5, p} : Finset ℕ).card = 2 := by
        rw [Finset.card_insert_of_notMem
          (by simp [Ne.symm hp_ne_5]), Finset.card_singleton]
      calc 2 = _ := h2.symm
        _ ≤ _ := Finset.card_le_card hsub
    exact multiple_factors_yields_Hd h5p_in_H hcard
  · exact hS_inf

/-- Theorem 3: the partition theorem. -/
theorem H_partition :
    H = Hd ∪ Hu ∧ Hd ∩ Hu = ∅ ∧ Hd.Infinite :=
  ⟨H_eq_Hd_union_Hu, Hd_inter_Hu, Hd_infinite⟩

end PrimPythHyp
\end{leancode}

Once (F), (G-easy), (G-uniq) are themselves formalized, replacing the
\verb|axiom| declarations with \verb|theorem| proofs, this file would
become a self-contained, machine-checked proof of Theorem 3 with no
classical inputs assumed beyond what is already in Mathlib.

\end{document}


