\input macro

\parindent=0pt
\parskip=.5\baselineskip
\baselineskip = 1.1\baselineskip

\footline{\hss\tenrm\folio\hss}

\mantitle{Properties of a Linear Unbiased Minimum Variance Estimator}
{R. Scott McIntire}{May 2, 2026}

\section{Overview}
A standard problem is to find the linear unbiased ``minimum'' variance estimator
for the overdetermined problem\footnote{\kern 0.5pt \raise 0.5ex \hbox{\dag}}
{This means that $n > m$.}: $A {\bf x} = {\bf b}$, where
$A \in {\bf R}^{n \times m}$, ${\bf b} \in {\bf R}^n$ is a random observation
with $E[{\bf b}] = A {\bf x}$ and covariance $V$ (assumed invertible), and
${\bf x} \in {\bf R}^m$ is the deterministic parameter we wish to estimate.
We look for a linear unbiased estimator for ${\bf x}$ of the form
${\bf x}^* = L {\bf b}$, where $L\in {\bf R}^{m \times n}$. Since ${\bf x}^*$
is unbiased we must have
$E[{\bf x} - {\bf x}^*] = {\bf x} - L A {\bf x} = (I - LA){\bf x}$. For this
to vanish for every ${\bf x}$, we require $LA = I$. This implies that $A$
and $L$ have rank $m$.
The covariance of $L{\bf b}$ is $LVL^T$. The problem now is to find
the best $L$ such that $LA = I$ and the covariance matrix, $LVL^T$, is
minimized in some sense.

The problem may be restated as choosing $L$ to ``minimize'' the norm of $LVL^T$
subject to the constraint $LA = I$. The question is what norm does one
use. The ``natural'' norm of a matrix is $\sqrt{{\rm tr}\left(M^TM\right)}$. This 
    follows from considering the standard basis in ${\bf R}^{n \times m}$ and
    using the usual norm of the vector space ${\bf R}^{n \times m}$. 
The matrix of interest, $LVL^T$, is positive definite
    symmetric; in this case, each of the ``trace''
    expressions, $\root k \of {{\rm  tr}\left(M^k\right)} \;\; (k \in [1,\infty ))$, is a norm for a
positive definite matrix $M_{m \times m}$ (these are the {\it Schatten\/} $k$-norms.)

\section{Do the Norms Matter?}
The question is, do different choices of norms lead to a different
solution, $L$? To answer this, we solve the constrained minimization
problem that results from each of these trace norms\footnote{\kern 0.5pt \raise 0.5ex \hbox{\ddag}}
{Equivalently, we minimize the $k^{\rm th}$ power
of the $k^{\rm th}$ trace norm.}:
$$
F(L; k) = {\rm tr}\left((LVL^T)^k - (LA - I)\Lambda_{k}^T\right) 
\quad k \in [1,\infty)
$$
where $\Lambda_k \in {\bf R}^{m \times m}$ is a Lagrange multiplier.

The derivative of $F$ with respect to $L$ is given by:
$$
\eqalignno{
DF(L)(H) & = {\rm tr}\left[ k(LVL^T)^{k-1}(HVL^T + LVH^T) - 
HA\Lambda_k^T\right]}
$$
At a minimum value, $L^*$, the derivative must vanish; therefore,
$$
\eqalignno{
& {\rm tr}\left[ k(L^*VL^{*T})^{k-1}(HVL^{*T} + L^*VH^T) - 
HA\Lambda_k^T\right] = 0
  \quad \forall H \in {\bf R}^{m \times n} & (1) \cr}
$$

Before going further we list some facts regarding the trace inner product that 
will be used in the rest of the paper.

\item{$\forall H \in {\bf R}^{n \times n},\; H \; \hbox{ anti-symmetric} \; \Rightarrow \; {\rm tr}(H) = 0$.}
\item{Given a matrix $A \in {\bf R}^{m \times n}$: $\forall H \in {\bf R}^{m \times n},\; {\rm tr}(A^T H) = 0 \; \Rightarrow \; A \equiv 0.$}
\item{$\forall A,H \in {\bf R}^{n \times n}, \;A \; {\rm symmetric} \; \wedge \; H \; \hbox{anti-symmetric}\;  \Rightarrow \; {\rm tr}(A^T H) = 0$.}
\item{Given a matrix $A \in {\bf R}^{n \times n}$: $\forall H \in {\bf R}^{n \times n}, \; H \; \hbox{anti-symmetric} \; \wedge \; {\rm tr}(A^T H) = 0 \; \Rightarrow A \; \hbox{is symmetric}$.}
\item{Given a matrix $A \in {\bf R}^{n \times n}$: $\forall H \in {\bf R}^{n \times n}, \; A \; \hbox{symmetric} \; \wedge \; H \; \hbox{symmetric} \; \wedge \; {\rm tr}(A^T H) = 0 \; \Rightarrow A \equiv 0$.}

The first fact follows from a property of square matrices; namely,
${\rm tr}(A) = {\rm tr}(A^T)$; and the definition of an anti-symmetric $A$, $A^T = -A$.
The second fact follows since any inner product on a vector space, $V$, satisfies: 
$\forall y \in V, \langle x, y \rangle = 0 \; \Rightarrow \; x \equiv 0$. 
The third fact follows -- using Fact 1 (or rather its underlying identity ${\rm tr}(M) = {\rm tr}(M^T)$) -- as follows.
If $S$ and $W$ are symmetric and anti-symmetric matrices in ${\bf R}^{n \times n}$, then
${\rm tr}(S \, W) = {\rm tr}((S \, W)^T) = {\rm tr}(W^T \, S^T) = -{\rm tr}(W \, S) = - {\rm tr}(S\, W)$.
This implies that ${\rm tr}(S \, W) = 0$.
The fourth fact follows from the unique decomposition of any matrix
$A \in {\bf R}^{n \times n}$ as a sum of a symmetric and an anti-symmetric matrix:
$A = A^{s} + A^{w}$, so $A^T = A^{s} - A^{w}$. For any anti-symmetric $H$,
${\rm tr}(A^T H) = {\rm tr}(A^s H) - {\rm tr}(A^w H) = -{\rm tr}(A^w H)$ by Fact 3.
So if ${\rm tr}(A^T H) = 0$ for all anti-symmetric $H$, then ${\rm tr}(A^w H) = 0$ for all
such $H$. Setting $H = A^w$ gives ${\rm tr}((A^w)^2) = -{\rm tr}((A^w)^T A^w) = -\|A^w\|_F^2 = 0$,
so $A^{w} \equiv 0$ and $A$ is symmetric.
The fifth fact follows by setting $H$ to $A$. Again, the fact that the trace inner product induces a norm implies 
that $A = 0$.


Continuing with the derivation, in the special case $k = 1$, equation (1) 
implies that 
$ 2VL^{*T}  = A\Lambda_1^T$; or, 
$$
\eqalignno{
2L^{*T} & = V^{-1} A \Lambda_1^T & (2) \cr }
$$
Multiplying both sides by $A^T$ and using the fact that
$A^TL^{*T} = I$ yields $2I = A^TV^{-1}A\Lambda_1^T$. Since $A$ has rank
$m$, $A^TV^{-1}A$ is invertible, so $\Lambda_1^T = 2(A^TV^{-1}A)^{-1}$
(the right-hand side is symmetric, so $\Lambda_1 = \Lambda_1^T$). Combining
this with equation (2) gives
$$
L^* = (A^TV^{-1}A)^{-1}A^TV^{-1}
$$

We now proceed with the case when $k > 1$. Given any ${\tilde H} \in {\bf R}^{m \times m}$, 
let $H = {\tilde H}A^TV^{-1}$. Then equation (1) becomes:

$$
\eqalignno{
& {\rm tr}\left[ k(L^*VL^{*T})^{k-1}({\tilde H}A^TL^{*T} + L^*A{\tilde H}^T) -
  {\tilde H}(A^TV^{-1}A)\Lambda_k^T\right]
= 0 \quad \forall {\tilde H} \in {\bf R}^{m \times m}  \cr }
$$

Or,
$$
\eqalignno{
& {\rm tr}\left[ k(L^*VL^{*T})^{k-1}({\tilde H} + {\tilde H}^T) -
  {\tilde H}(A^TV^{-1}A)\Lambda_k^T\right]
= 0 \quad \forall {\tilde H} \in {\bf R}^{m \times m} & (3)  \cr }
$$
since $L^*A = I$ and $A^TL^{*T} = I$.


Equation (3) holds for all ${\tilde H} \in {\bf R}^{m \times m}$. Let
$M = (L^*VL^{*T})^{k-1}$ and $C = A^TV^{-1}A$. Since $M$ is symmetric,
${\rm tr}(M(\tilde H + \tilde H^T)) = 2\,{\rm tr}(\tilde H M)$ by cyclic
invariance, so (3) reads
$$
{\rm tr}\bigl[\tilde H (2kM - C\Lambda_k^T)\bigr] = 0
\quad \forall {\tilde H} \in {\bf R}^{m \times m}.
$$
By Fact 2, $2kM - C\Lambda_k^T = 0$. Since $A$ has rank $m$,
$(A^TV^{-1}A)^{-1}$ exists, and we can solve for $\Lambda_k^T$:
$$
\Lambda_k^T = 2k\,(A^TV^{-1}A)^{-1}(L^*VL^{*T})^{k-1}.
$$

Equation (1) now becomes:

$$
{\rm tr}\left[k(L^*VL^{*T})^{k-1}(HVL^{*T} + L^*VH^T) -
  2kHA(A^TV^{-1}A)^{-1}(L^*VL^{*T})^{k-1}\right] = 0 \quad 
\forall H \in {\bf R}^{m \times n}
$$

Or,

$$
{\rm tr}\left[\biggl( (HVL^{*T} + L^*VH^T) / 2 -
  HA(A^TV^{-1}A)^{-1}\biggr) (L^*VL^{*T})^{k-1}\right] = 0 \quad 
  \forall H \in {\bf R}^{m \times n}
$$

Which may be written
$$
{\rm tr}\left[\biggl( HVL^{*T} + ( L^*VH^T  - HVL^{*T} )/ 2 -
  HA(A^TV^{-1}A)^{-1}\biggr) (L^*VL^{*T})^{k-1}\right] = 0 \quad 
  \forall H \in {\bf R}^{m \times n}
$$

Since $( L^*VH^T  - HVL^{*T} )/ 2$ is anti-symmetric and $(L^*VL^{*T})^{k-1}$
is symmetric, the trace of their product is zero. Therefore we may write
the last equation as

$$
 {\rm tr}\left[\biggl( HVL^{*T} -
  HA(A^TV^{-1}A)^{-1}\biggr) (L^*VL^{*T})^{k-1}\right] = 0 \quad \forall H
    \in {\bf R}^{m \times n}
$$

Or,

$$
 {\rm tr}\left[H\biggl( VL^{*T} -
  A(A^TV^{-1}A)^{-1}\biggr) (L^*VL^{*T})^{k-1}\right] = 0 \quad \forall H
    \in {\bf R}^{m \times n} 
$$

This implies that $\left( VL^{*T} - A(A^TV^{-1}A)^{-1}\right)
(L^*VL^{*T})^{k-1} = 0 $.
Since $L^*$ has rank $m$, $(L^*VL^{*T})^{-(k-1)}$ exists, so that
$ VL^{*T} - A(A^TV^{-1}A)^{-1} = 0 $. Hence, 
$$
L^* = (A^TV^{-1}A)^{-1}A^TV^{-1}
$$


Therefore, it does not matter which of the matrix norms we
use: the ``best'' linear unbiased estimator is the same.

\vskip .25in
\section{Best Estimator under the Spectral Norm}
We now use the results of the trace norms above to show that 
the estimator obtained has the smallest maximum eigenvalue 
over all linear estimators.

\proclaim Lemma 1. If $\Omega$ is a closed set in ${\bf R}^m$ and if
$x^* \in \Omega$ is a {\it common\/} solution to
$\min_{\{x \in \Omega: g(x) = 0\}}\;\varphi_k(x)$
for {\it every\/} $k \in [1,\infty)$; further,
if  $\varphi_k$ are functions which converge point-wise to the
continuous function $\varphi$ on $\Omega$ with the coercive property
$\varphi(x) \to \infty$ as $\|x\| \to \infty$, then
$x^* \in \mathop{\rm argmin}_{\{x \in \Omega: g(x) = 0\}} \varphi(x).$

{\bf Proof.} 

{\it Existence:\/}
The set $\{x \in \Omega: g(x) = 0\}$ is non-empty.
Let $\{x_n\}$ be a minimizing sequence in this set; that is, $\varphi(x_n)$
converges to the infimum.
By the coercive property of $\varphi$, $\{x_n\}$ is bounded.
Since $\Omega$ is closed, a subsequence converges to some
$x^\prime \in \Omega$. By the continuity of $\varphi$ on $\Omega$,
$\varphi(x^\prime)$ equals the infimum, which is therefore finite and attained.

We proceed to show that $x^*$ is a solution.
Let ${\tilde x}$ be a solution of the problem 
$\min_{\{x \in \Omega : g(x) = 0\}}\; \varphi(x)$;
then for any $\epsilon > 0$, we have, for sufficiently large $n$, 
(since $\varphi_n$ converges point-wise to $\varphi$)
$$
 \varphi_n({\tilde x}) < \varphi({\tilde x}) + \epsilon
$$
Since $x^*$ is a solution for $\varphi_n$, we also have
$$
\varphi_n(x^*) \le \varphi_n({\tilde x})
$$
Combining the two gives

$$
\varphi_n(x^*) \le \varphi_n({\tilde x}) < \varphi({\tilde x}) + \epsilon
$$

Or,
$$
\varphi_n(x^*) \le \varphi({\tilde x}) + \epsilon
$$

Taking the limit as $n \rightarrow \infty$ we have:

$$
\varphi(x^*) \le \varphi({\tilde x}) + \epsilon
$$

Since this is true for all $\epsilon > 0$ we necessarily have that
$$
\varphi(x^*) \le \varphi({\tilde x})
$$

Noting that $x^*$ satisfies the constraint, $g(x) = 0$, we infer that 
$x^*$ is also a solution.


\proclaim Theorem 1. The linear unbiased estimator previously obtained for the
least squares problem, $L^*{\bf b} =
(A^TV^{-1}A)^{-1}A^TV^{-1}{\bf b}$, is also the solution that
minimizes the maximum eigenvalue of the estimator's covariance matrix, 
$LVL^T$, subject to the constraint $L_{m \times n}A_{n \times m} = I_{m \times m}$.

{\bf Proof.} We apply Lemma 1 with the variable $L \in {\bf R}^{m \times n}$,
$\Omega = \{L \in {\bf R}^{m \times n} : LA = I\}$ (closed),
$\varphi_n(L) = \root n \of {{\rm tr}((LVL^T)^n)}$ (the $n^{\rm th}$ Schatten norm of $LVL^T$),
and $\varphi(L) = \lambda_{\max}(LVL^T)$ (the spectral norm of $LVL^T$, which is
positive definite since $L$ has full row rank and $V$ is positive definite).
The constraint $g(L) = 0$ is vacuous since $\Omega$ already encodes $LA = I$.
Theorem 1 follows after collecting the facts:

\item{For a positive definite matrix with eigenvalues $\lambda_i > 0$,
$\root k \of {\sum_i \lambda_i^k} \to \max_i \lambda_i$ as $k \to \infty$.}
\item{The $L_2$ matrix norm of a positive definite matrix is its
maximum eigenvalue.}
\item{The $L_2$ norm is coercive.}


\bye
