\documentclass{article}
\usepackage{amsmath,amscd}

\makeatletter
\newcommand{\specialcell}[1]{\ifmeasuring@#1\else\omit$\displaystyle#1$\ignorespaces\fi}
\makeatother

\title{Origins of Matrix/Vector and Matrix/Matrix Multiplication}
\author{R. Scott McIntire}
\date{Apr 25, 2026}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{7.5in}
\setlength{\hoffset}{-0.75in}
\setlength{\voffset}{-0.75in}

\setlength{\parindent}{0.0in}
\setlength{\parskip}{0.5\baselineskip}


\begin{document}
\maketitle


\section{Generalizing $a \, x = b$}
Why is matrix/vector and matrix/matrix multiplication defined the way it is?
One motivation is to try to extend the (albeit trivial) solution of linear
equations in 1-dimension to n dimensions. To do this, we go through the solution
of the 1-dimensional case in careful detail.

The scalar problem is:
\begin{align}
  a \, x &= b \label{scalar-problem}
\end{align}

Here is a very explicit solution keeping an eye towards generalization.
\begin{align}
  a \, x &= b \\
  a^{-1} ( a \, x ) &= a^{-1} b \quad \quad \text{(Multiply by Inverse)} \\
  (a^{-1} a) x &= a^{-1} b \quad \quad \text{(Use Associativity of multiplication)} \\
  1 \, x &= a^{-1} b \quad \quad \text{(What Inverse multiplication does)} \\
  x &= a^{-1} b \quad \quad \text{(What Identity multiplication does)}
\end{align}

The multi-dimensional problem has a lot more variables and coefficients.
To start, we need to be more systematic about the naming of these coefficients.
With this in mind, the multi-dimensional linear problem can be written:
\begin{align*}
  a_{11} x_1 + a_{12} x_2 + \cdots + a_{1n} x_n &= b_1 \\
  a_{21} x_1 + a_{22} x_2 + \cdots + a_{2n} x_n &= b_2 \\
  \specialcell{\hfil\cdots\hfil} &= \vdots \\
  a_{n1} x_1 + a_{n2} x_2 + \cdots + a_{nn} x_n &= b_n 
\end{align*}
Here, $a_{ij}$ is the coefficient in the $i^{\rm th}$ row and $j^{\rm th}$ column.


To make this look like the 1-dimensional case, we need to think of the $b$'s as
a single unit. Our single unit will be the {\em vector\/} of the $b$'s.

The multi-dimensional case can be rewritten in vector terms as:
\begin{align}
  \left[
	  \begin{array}{c}
	  a_{11} x_1 + a_{12} x_2 + \cdots + a_{1n} x_n  \\
	  a_{21} x_1 + a_{22} x_2 + \cdots + a_{2n} x_n \\
  \specialcell{\hfil\vdots\hfil}  \\
  a_{n1} x_1 + a_{n2} x_2 + \cdots + a_{nn} x_n 
  \end{array}
  \right] &= \left[
                 \begin{array}{c}
                   b_1 \\
                   b_2 \\
                   \vdots \\
                   b_n
                 \end{array}
   \right] \label{vector-form}
\end{align}
We can replace the tedious list of sums on the left hand side of the above
with standard mathematical summing notation:
\begin{align}
  \left[
  \begin{array}{c}
	\sum_{j=1}^n a_{1j} x_j        \\ 
	\sum_{j=1}^n a_{2j} x_j        \\ 
  \specialcell{\hfill\vdots\hfill} \\
	\sum_{j=1}^n a_{nj} x_j        \\ 
  \end{array}
  \right] &= \left[
                 \begin{array}{c}
                   b_1     \\
                   b_2     \\
                   \vdots  \\
                   b_n
                 \end{array}
   \right] \label{vector-form1}
\end{align}
Two things need to be done: treat the $x$'s as a unit -- as we did with the
$b$'s -- and separate the $a$ coefficients from the $x$'s. This must be done formally
and yet have the same meaning as the original problem formulation.
This is done in the most natural of ways:
\begin{align}
  \left[
  \begin{array}{cccc}
    a_{11} & a_{12} & \cdots & a_{1n}     \\
	a_{21} & a_{22} & \cdots & a_{2n}     \\
	&& \specialcell{\hfill\vdots\hfill} & \\
    a_{n1} & a_{n2} & \cdots & a_{nn}     \\
    \end{array}
  \right]
  \left[
  \begin{array}{c}
    x_1    \\
    x_2    \\
    \vdots \\
    x_n
  \end{array}
  \right] &=
                \left[
                \begin{array}{c}
                  b_1    \\
                  b_2    \\
                  \vdots \\
                  b_n
                \end{array}
  \right] \label{matrix-form2}
\end{align}
Taking the rectangular collection (matrix) of $a$'s to be represented by 
$A$; the (vector) collection
of $x$'s to be ${\bf x}$; and the collection of the $b$'s to be ${\bf b}$;
we may write \eqref{matrix-form2} as:%
\footnote{In general, we use uppercase italic letters for matrices; lowercase bold roman letters
for vectors; and lowercase italic letters for scalars.}
\begin{align}
  A \, {\bf x} &= {\bf b} \label{concise-matrix-form}
\end{align}
Here, $A$ represents the ``matrix'' of the $a$'s, with $A_{i,j} = a_{ij}$; meaning, that the
$i^{\rm th}$ row and $j^{\rm th}$ column of $A$ is $a_{ij}$.
This now looks like the scalar problem, \eqref{scalar-problem}.
Just as one writes $a \, x = b$, interpreting the proximity of $a$ and $x$ to mean
multiplication, we want the proximity of $A$ and ${\bf x}$ to mean matrix/vector
multiplication -- something we need to work out.

\section{Solution of $A {\bf x} = {\bf b}$ determines Matrix Multiplication}

We proceed to try solving \eqref{concise-matrix-form} using the solution 
procedure used above for the scalar case. In the process, we need to:
\begin{itemize}
    \item Vectorize the input variable $x$ and the outputs $b$. -- {\bf Done}.
    \item Define an ``$a$'' matrix. -- {\bf Done}.
    \item Define matrix-vector multiplication.
    \item Define matrix-matrix multiplication.
    \item Define the identity matrix.
    \item Define the inverse matrix.
\end{itemize}
Here, in our new notation, is what the solution outline looks like for the multi-dimensional problem:
\begin{align}
	A \, {\bf x} &= {\bf b} \label{concise-matrix-form2}\\
	A^{-1} ( A \, {\bf x} ) &= A^{-1} {\bf b} \quad \quad \text{(Matrix/vector multiply the vector equation by $A$ Inverse)} \label{inv-vec} \\
	(A^{-1} A) {\bf x} &= A^{-1} {\bf b} \quad \quad \text{(Use Associativity of matrix/matrix multiplication)} \label{assoc-vec} \\
	I \, {\bf x} &= A^{-1} {\bf b} \quad \quad \text{(What Inverse matrix/matrix multiplication does)} \\
	{\bf x} &= A^{-1} {\bf b} \quad \quad \text{(What Identity matrix/vector multiplication does)}
\end{align}
How do we make sense of this matrix/vector syntax on the left hand side of 
\eqref{concise-matrix-form2}? It should have
the proper meaning; that is, \eqref{matrix-form2}
should have the same meaning as \eqref{vector-form1}.

Examining \eqref{matrix-form2} and the left hand side of
\eqref{vector-form1}, gives us our definition of matrix/vector multiplication:
\begin{align}
  \left[A \, {\bf x}\right]_i &\equiv \sum_{j=1}^n A_{ij} x_j
                                         \label{matrix-vector-mult}
\end{align}
That is, $A$ acts on a vector, ${\bf x}$, to create a new vector, $A \, {\bf x}$,
whose $i^{\rm th}$ entry is the $i^{\rm th}$ entry of
the left hand side of \eqref{vector-form1}.

Another way to write this is:
\begin{align}
	A {\bf x} &= \sum_{j=1}^n x_j {\bf A}^j \label{mult-as-vec-combo}
\end{align}
Here, ${\bf A}^j$ is the $j^{\rm th}$ column vector of $A$. 

This can be seen by writing \eqref{vector-form} as:%
\footnote{Here, we assume the reader has a passing knowledge of how vectors 
are added and multiplied by scalars.}
\begin{align}
  \begin{array}{ccc}
	  x_1 \left[ \begin{array}{c}
			  a_{11} \\
			  a_{21} \\
		  	  \vdots \\
			  a_{n1}
		     \end{array}
	      \right]
		  +  
	  x_2 \left[ \begin{array}{c}
			  a_{12} \\
			  a_{22} \\
		  	  \vdots \\
			  a_{n2}
		     \end{array}
	      \right]
		  + \; \cdots \;
		  + 
	  x_n \left[ \begin{array}{c}
			  a_{1n} \\
			  a_{2n} \\
		  	  \vdots \\
			  a_{nn}
		     \end{array}
	      \right]
	  & = &
                \left[
                \begin{array}{c}
                  b_1    \\
                  b_2    \\
                  \vdots \\
                  b_n
                \end{array}
  \right]
  \end{array}
\end{align}

From this we can see that matrix/vector multiplication of $A$ on ${\bf x}$ is
a ``weighted'' combination of the columns of $A$ using the elements of ${\bf x}$
as the weights. Notice that if we pick an ``${\bf x}$'' with all of the weight 
concentrated on the $j^{\rm th}$ entry, then matrix/vector multiplication on this 
specific vector yields the $j^{\rm th}$ 
column of $A$, ${\bf A}^j$. Specifically, applying $A$ to the special
vectors, ${\bf e}_j$
-- which are $0$ everywhere except at $j$ where they are $1$ --
we see that $A \, {\bf e}_j = {\bf A}^j$. 
Consequently, matrix/vector multiplication on this set of vectors
uniquely determines $A$
-- if there is another matrix that gives the same result for these vectors with
respect to matrix/vector multiplication, 
its columns would have to match $A$.

We note the following for future reference.
\begin{align}
	[{\bf A}^j]_i = A_{i, j} \label{vec-matrix-index}
\end{align}
That is, the $i^{\rm th}$ entry of the column vector, ${\bf A}^j$, is the
$i^{\rm th}$ row,  $j^{\rm th}$ column of the matrix $A$. 

Before moving on with our outline, we note one very important property of 
matrix/vector multiplication: It is {\em linear\/}.
By that we mean the following:%
\footnote{You might ask: ``How did someone think up such a 
strange property?'' But it is not strange, it is what one might hope for, as all we
are saying is that matrix/vector multiplication works the same way that ordinary {\em scalar\/} 
multiplication does -- it {\em distributes\/} through addition and scalar multiplication.} 

\begin{align}
	A( \alpha {\bf x} + \beta {\bf y} ) &= \alpha\, A{\bf x} + \beta\, A{\bf y}
\end{align}
One can show this from our definition of matrix/vector multiplication which we
leave to the reader. What this means in practice is that for a given vector, ${\bf z} = \alpha{\bf x} + \beta{\bf y}$,
we can compute $A {\bf z}$ by computing $A$ on the ``components'' of ${\bf z}$ and then multiplying by scalars and adding
the resulting vectors.
This can be an easier way to compute the {\em action\/} of a matrix on a vector in some cases.
The linearity result above can be extended to arbitrary sums as:
\begin{align}
	A\left( \sum_{j=1}^n c_j {\bf x}_j \right) &= \sum_{j=1}^n c_j A {\bf x}_j
\end{align}

Continuing with our solution outline, matrix/matrix multiplication is needed
to go from equation \eqref{inv-vec} to \eqref{assoc-vec}. How must this be defined?
Well, we need to make sense of:
\begin{align}
  (A^{-1} A) {\bf x} &= A^{-1} (A \, {\bf x}) 
\end{align}
This involves the inverse of the matrix $A$, which we also need to define. 
We want associativity to hold {\em for all\/} matrix products, not just $A^{-1} A$.
So we forgo what the inverse of a matrix might mean and focus on imposing the
condition that matrix/matrix multiplication be associative for {\em all\/} input vectors.
Specifically, the condition we impose is:
\begin{align}
	(B\, A) {\bf x} &= B (A \, {\bf x}) \quad \forall {\bf x} \in {\bf R}^n \label{ASSOC}
\end{align}
Notice that we are imposing what matrix/matrix multiplication is by saying how the 
new matrix formed, $B \, A$, acts on an {\em arbitrary\/} vector, ${\bf x}$. And we 
specify that action by the right hand side of \eqref{ASSOC}, which involves only matrix/vector
multiplication -- something we have defined.

We want to emphasize that this is a strong condition to impose; meaning, that the 
requirement that this is true for {\em all\/} ${\bf x}$ gives us a lot to work with.
For instance, if one had two $n\times n$ matrices, $C$ and $D$, what could you conclude if someone
told you that $C {\bf x} = D {\bf x}$ for {\em some\/} $n$ vector, ${\bf x}$?
Well, the answer is: not much.
However, if I told you that $C {\bf x} = D {\bf x}$ {\em for all\/} ${\bf x}$, what could you say?
I claim that tells us that $C \equiv D$; meaning that every entry in $C$ is the same as the corresponding
entry in $D$. We can see this by applying $C$ and $D$ to each of the ${\bf e}_j$, ($j \in [1, n]$) vectors.
For each $j$, $C {\bf e}_j$ and $D {\bf e}_j$ pick off the $j^{\rm th}$ column of 
$C$ and $D$ respectively. So we see that each column of $C$ and $D$ must match; consequently,
$C \equiv D$.

We apply this idea to \eqref{ASSOC} to find a formula for matrix/matrix multiplication.
Since \eqref{ASSOC} must be true {\em for all\/} ${\bf x}$; in particular, it must be
true for the vectors $\{{\bf e}_j\}_{j=1}^n$.
This gives us the following equations:
\begin{align}
	(B \, A) {\bf e}_j &= B (A \, {\bf e}_j) \quad\quad \text{$j \in [1,n]$} 
\end{align}
Or,
\begin{align}
	(B \, A)^j &= B \, {\bf A}^j \quad\quad \text{$j \in [1, n]$}
\end{align}
For any given, $i$, the $i^{\rm th}$ entry of the left and right hand side is
(after using the definition of matrix/vector 
multiplication, \eqref{matrix-vector-mult}, on the right hand side)
\begin{align}
	\left[(B \, A)^j\right]_i &= \sum_{k=1}^n B_{i,k} \left[{\bf A}^j\right]_k \quad \forall i, j \in [1, n]
\end{align}
Using \eqref{vec-matrix-index} on the left and right hand sides of the previous equation we have
\begin{align}
	(B \, A)_{i, j} &= \sum_{k=1}^n B_{i, k} A_{k, j} \quad \forall i, j \in [1, n]
\end{align}
And we are done: we have shown what the new matrix $B \, A$ is by giving every entry $(B\, A)_{i, j}$.
So using the set of vectors, $\{{\bf e}_j\}_{j=1}^n$, we have completely determined what 
the matrix multiplication of $B \, A$ is in order that matrix/matrix multiplication be 
associative when using these vectors. Does this mean that associativity works for 
{\em arbitrary\/} vectors? The answer is yes, by the linearity of
matrix/vector multiplication. Any given ${\bf x}$ can be written as a {\em linear combination\/}%
\footnote{Again, we assume the reader is familiar with expanding a vector in terms of the standard basis vectors.}
of the vectors $\{{\bf e}_j\}_{j=1}^n$: ${\bf x} = a_1 {\bf e}_1 + a_2 {\bf e}_2 + \cdots + a_n {\bf e}_n$.
Matrix multiply the left and right hand sides by the matrix, $B\,A$. This gives
\begin{align}
	(B\, A){\bf x} &= (B \, A)\left(\sum_{j=1}^n a_j {\bf e}_j\right) \quad\quad \text{Replace ${\bf x}$ with its components, ${\bf e}_j$.}\nonumber \\
				   &=\sum_{j=1}^n a_j (B \, A){\bf e}_j \quad\quad \text{{\em Linearity\/} of matrix/vector multiplication by $(B \, A)$.}\nonumber \\
				   &=\sum_{j=1}^n a_j B (A{\bf e}_j) \quad\quad \text{{\em Associativity\/} of matrix/matrix multiplication on components, ${\bf e}_j$.}\nonumber \\
				   &=B\left(\sum_{j=1}^n a_j A{\bf e}_j\right) \quad\quad \text{{\em Linearity\/} of matrix/vector multiplication by $B$.}\nonumber \\
				   &=B\left( A\left(\sum_{j=1}^n a_j {\bf e}_j\right)\right) \quad\quad \text{{\em Linearity\/} of matrix/vector multiplication by  $A$.}\nonumber \\
				   &=B( A {\bf x}) \quad\quad \text{Replace components with ${\bf x}$.}
\end{align}
In the above calculations, we applied our new matrix, $(B \, A)$ to an arbitrary ${\bf x}$, and 
used the {\em linearity\/} of matrix/vector multiplication and the {\em associativity\/}
of matrix/matrix multiplication on the components, $\{{\bf e}_j\}_{j=1}^n$, to show that matrix/matrix
multiplication, as we have defined it, is associative -- {\em irrespective\/} of the input vector.

The next order of business is to identify an $n \times n$ matrix which serves as an
identity (in terms of matrix/vector multiplication).
Let us suppose that we have such a matrix and let's call it $I$.
Then we must have $I \, {\bf x} = {\bf x} \quad \forall {\bf x} \in {\bf R}^n$. 
Then, in particular,  
$I \, {\bf e}_{j} = {\bf e}_j \quad \forall j \in [1, n]$. 
However, we know that $I \, {\bf e}_j = {\bf I}^j$. Therefore, $I$
must have the property that ${\bf I}^j = {\bf e}_j$. Consequently, $I$ must have the form:
\begin{align}
  I &= \left[
      \begin{array}{ccccc}
        1 & 0 & 0  & \cdots & 0 \\
        0 & 1 & 0  & \cdots & 0 \\
        0 & 0 & 1  & \cdots & 0 \\
		  &&&\specialcell{\hfil\ddots\hfil}  & \vdots  \\
        0 & 0 & 0  & \cdots & 1 \\
      \end{array}
  \right] \label{identity-matrix}
\end{align}
We have shown that the only candidate for the identity matrix with respect to 
matrix/vector multiplication is the matrix $I$. That is, if there is an 
identity matrix, it must be the matrix of equation \eqref{identity-matrix}. Does it satisfy
the property of being an identity matrix (again, in the matrix/vector multiplication world)?
That is, do we have $I \, {\bf x} = {\bf x}$ for all ${\bf x} \in {\bf R}^n$?
As before, we can express ${\bf x}$ as: ${\bf x} = \sum_{j=1}^n c_j {\bf e}_j$.
By the {\em linearity\/} of matrix/vector multiplication, we have
\begin{align}
	I {\bf x} &= I\left( \sum_{j=1}^n c_j {\bf e}_j \right)  \quad\quad \text{Replace ${\bf x}$ with components, ${\bf e}_j$.} \nonumber \\
							 &= \sum_{j=1}^n c_j  I{\bf e}_j \quad\quad \text{{\em Linearity\/} of matrix/vector multiplication.} \nonumber \\
							 &= \sum_{j=1}^n c_j {\bf e}_j \quad\quad \text{{\em Identity\/} of $I$ on components, ${\bf e}_j$.}\nonumber \\
							 &= {\bf x} \quad\quad \text{Replace components with ${\bf x}$.}
\end{align}
One can show that this identity matrix, $I$, is also the identity operator for
matrix/matrix multiplication.

The only thing left is to know when a matrix inverse exists and how to compute it.
We do not attempt to do this in this paper. In the next section we continue with
a qualitative comparison of the solution to the scalar problem, $a\, x = b$, 
and its vectorized cousin.


\section{Qualitative Features of Solutions}
We can view the multiplication of two numbers, $a$ and $x$, as just that.
Or, we can think of $a$ being fixed and letting $x$ ``run-through'' all 
numbers. Here we see two cases: if $a \neq 0$, then letting $x$ run
through all of the numbers in ${\bf R}$ will produce all of the numbers in ${\bf R}$.
We could think of $a$ as an ``operator'' and call the set of all possible 
outputs, the range of $a$ and denote it: ${\cal R}(a)$.
There is another case, $a$ could be zero. In this case it's range is the
set $\{0\}$. In the first case, with non-zero $a$ it is clear that we can
find an $x$ to ``hit'' a given value $b$. That is, we can solve $a\, x = b$.

One can define the same concept for a matrix, $A$. In this case, we produce a 
``space'' of all possible outputs of $A$ ``acting'' using all possible input vectors; 
that is, all $x \in {\bf R}^n$. This amounts to, using \eqref{mult-as-vec-combo}, the set of all 
{\em linear combinations\/} of the columns of $A$. Using this language
of ranges, here is what we can say about the scalar problem: $a \, x = b$.
\begin{description}
	\item[Unique Solution:] If $a^{-1}$ exists ($a \neq 0$),
		then there is a {\bf unique\/} solution.
	\item[No Solution:] If $a^{-1}$ does not exist (i.e., $a = 0$) {\em AND\/}
		$b$ is {\bf not\/} in the range of $a$ (that is: $b \ne 0$), then there is {\bf no\/} solution.
	\item[Infinite Solutions:] If $a^{-1}$ does not exist (i.e., $a = 0$) {\em BUT\/}
		$b$ is in the range of $a$ (that is: $b = 0$),
		then there are an {\bf infinite\/} number of solutions.
\end{description}

Here is the analog of this solution categorization for the multi-dimensional case: $A \, {\bf x} = {\bf b}$.
\begin{description}
	\item[Unique Solution:] If $A^{-1}$ exists,
		then there is a {\bf unique\/} solution.
	\item[No Solution:] If $A^{-1}$ does not exist {\em AND\/}
		${\bf b}$ is not in the range of $A$ (that is ${\bf b} \notin {\cal R}(A)$),
		then there is {\bf no\/} solution.
	\item[Infinite Solutions:] If $A^{-1}$ does not exist {\em BUT\/}
		${\bf b}$ is in the range of $A$ (that is: ${\bf b} \in {\cal R}(A)$),
		then there are an {\bf infinite\/} number of solutions.
\end{description}

\end{document}

